---
title: "Data visualization I"
author: ""
date: ""
output: html_document
editor_options: 
  chunk_output_type: console
---

# Main Ideas

- Data visualization is an **extremely** effective way to express information 
  and extract meaning from data.
- We can build up an effective visualization systematically, layer by layer, 
  using a grammar of graphics (`ggplot2`).

"The simple graph has brought more information to the data analyst's mind than 
any other device" - John Tukey

# Lecture notes and exercises

Before we start, we need to configure git so that RStudio can 
communicate with GitHub. This requires two pieces of information: your email
address and your GitHub username.

Type the following lines of code in the **console** in RStudio, filling in your
username and the email address associated with your GitHub account.

```{r eval = FALSE}
library(usethis)
use_git_config(user.name= "github username", user.email="your email")
```

Next, load the `tidyverse` package. Recall, a package is just a bundle of 
shareable code.

```{r message = FALSE}
library(tidyverse)
```

Exploratory data analysis (EDA) is an approach to analyzing datasets in order to
summarize the main characteristics, often with visual representations of the 
data. We can also calculate summary statistics and perform data 
wrangling, manipulation, and transformation.

We will use `ggplot2` to construct visualizations. The gg in `ggplot2` stands 
for "grammar of graphics", a system or framework that allows us to describe the 
components of a graphic, building up an effective visualization layer by layer.

## Minneapolis housing data

We will introduce visualization using data on single-family homes sold in 
Minneapolis, Minnesota between 2005 and 2015.

**Question:** What happens when you click the green arrow in the code chunk
below? What changes in the "Environment" pane?

```{r load-data, message = FALSE}
mn_homes <- read_csv("data/mn_homes.csv")
```

```{r glimpse-data}
glimpse(mn_homes)
```

**Question:** What does each row represent? Each column?

Each row is a single-family home sold in Minneapolis, Minnesota; each column 
is a variable or feature related to the home.

### First visualization

`ggplot()` creates the initial base coordinate system that we will add layers 
to. We first specify the dataset we will use with `data = mn_homes`. The 
`mapping` argument is paired with an aesthetic (`aes`), which tells us how the 
variables in our dataset should be mapped to the visual properties of the graph.

**Question:** What does the code chunk below do?

```{r base-layer}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice))
```

```{r add-points}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice)) + 
   geom_point()
```

```{r add-smooth}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice)) + 
   geom_point() + 
   geom_smooth()
```

Run `?geom_smooth` in the console. What does this function do?

```{r add-labels}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice)) + 
   geom_point() + 
   geom_smooth() +
   labs(title = "Sales price vs. area of homes in Minneapolis, MN",
        x = "Area (square feet)", y = "Sales Price (dollars)")
```

The procedure used to construct plots can be summarized using the code below.

```{r eval = FALSE}
ggplot(data = [dataset], 
       mapping = aes(x = [x-variable], y = [y-variable])) +
   geom_xxx() +
   geom_xxx() + 
   other options
```

**Question:** What do you think `eval = FALSE` is doing in the code chunk above?

Chunk option `eval = FALSE` will include the code in the knitted document, but
the code will not be evaluated.

## Aesthetics

An aesthetic is a visual property of one of the objects in your plot.

- shape
- color
- size
- alpha (transparency)

We can map a variable in our dataset to a color, a size, a transparency, and 
so on.

**Question:** What will the visualization look like below? Write your answer 
down before running the code.

```{r aesthetic-1}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice,
                     color = fireplace)) + 
   geom_point() + 
   labs(title = "Sales price vs. area of homes in Minneapolis, MN",
        x = "Area (square feet)", y = "Sales Price (dollars)")
```

**Question:** What about this one?

```{r aesthetic-2}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice,
                     shape = fireplace)) + 
   geom_point() + 
   labs(title = "Sales price vs. area of homes in Minneapolis, MN",
        x = "Area (square feet)", y = "Sales Price (dollars)")
```

**Question:** This one?

```{r aesthetic-3}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice,
                     color = fireplace,
                     size = lotsize)) + 
   geom_point() + 
   labs(title = "Sales price vs. area of homes in Minneapolis, MN",
        x = "Area (square feet)", y = "Sales Price (dollars)")
```

**Question:** Are the above visualizations effective? Why or why not? How might
you improve them?

Color is one of the most effective tools at highlighting differences. Shape and
size and much more difficult to differentiate. The guide at
https://github.com/GraphicsPrinciples/CheatSheet/blob/master/NVSCheatSheet.pdf
discusses many of these considerations and what makes a good visualization.

**Question:** What is the difference between the two plots below?

The location of `color = "blue"`.

```{r compare-plots}
ggplot(data = mn_homes) + 
  geom_point(mapping = aes(x = area, y = salesprice, color = "blue"))

ggplot(data = mn_homes) + 
  geom_point(mapping = aes(x = area, y = salesprice), color = "blue")
```

Use `aes` to map variables to plot features, use arguments in `geom_xxx` for
customization not mapped to a variable.

Mapping in the `ggplot` function is global, meaning they apply to every layer
we add. Mapping in a particular `geom_xxx` function treats the mappings as 
local.

**Question:** Create a scatterplot using variables of your choosing using the 
`mn_homes` data.

```{r scatterplot}
ggplot(mn_homes, mapping = aes(x = lotsize, y = salesprice)) +
   geom_point() +
   labs(x = "Lot size", y = "Selling price")
```

**Question:** Modify your scatterplot above by coloring the points for each
community.

```{r scatterplot-community}
ggplot(mn_homes, mapping = aes(x = lotsize, y = salesprice, color = community)) +
   geom_point() +
   labs(x = "Lot size", y = "Selling price", color = "Community")
```

## Faceting

We can use smaller plots to display different subsets of the data using 
faceting. This is helpful to examine conditional relationships.

Let's try a few simple examples of faceting. Note that these plots should be 
improved by careful consideration of labels, aesthetics, etc.

```{r facet-1}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice)) + 
   geom_point() + 
   labs(title = "Sales price vs. area of homes in Minneapolis, MN",
        x = "Area (square feet)", y = "Sales Price (dollars)") + 
   facet_grid(. ~ beds)
```

```{r facet-2}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice)) + 
   geom_point() + 
   labs(title = "Sales price vs. area of homes in Minneapolis, MN",
        x = "Area (square feet)", y = "Sales Price (dollars)") + 
   facet_grid(beds ~ .)
```

```{r facet-3}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice)) + 
   geom_point() + 
   labs(title = "Sales price vs. area of homes in Minneapolis, MN",
        x = "Area (square feet)", y = "Sales Price (dollars)") + 
   facet_grid(beds ~ baths)
```

```{r facet-4}
ggplot(data = mn_homes, 
       mapping = aes(x = area, y = salesprice)) + 
   geom_point() + 
   labs(title = "Sales price vs. area of homes in Minneapolis, MN",
        x = "Area (square feet)", y = "Sales Price (dollars)") + 
   facet_wrap(~ community)
```

`facet_grid()`

- 2d grid
- rows ~ cols
- use . for no plot

`facet_wrap()`

- 1d ribbon wrapped into 2d

## Practice

(1) Modify the code outline to make the changes described below.

      - Change the color of the points to green.
      - Change `alpha` to make the points more transparent.
      - Add labels for the x axis, y axis.
      - Add an informative title.

When you are finished, remove `eval = FALSE` and knit the file to see the 
changes.

```{r modify-scatterplot, eval = FALSE}
ggplot(data = mn_homes, 
       mapping = aes(x = lotsize, y = salesprice)) + 
   geom_point(color = "darkgreen", alpha = 0.5) + 
   labs(title = "Sales price vs. lot size", 
        x = "Lot size", y = "Sale price ($)")
```

(2) Modify the code outline to make the changes described below.

      - Create a histogram of `lotsize`.
      - Modify the histogram by adding `fill = "green"` inside the 
        `geom_histogram()` function. 
      - Modify the histogram by adding `color = "red"` inside the 
        `geom_histogram()` function.

When you are finished, remove `eval = FALSE` and knit the file to see the 
changes. What is the difference between the `color` and `fill` arguments?

`fill` colors the area of the object, `color` colors the border of the
rectangles. For geometries that have an area, this is how `fill` and `color`
function.

```{r modify-histogram, eval = FALSE}
ggplot(data = mn_homes, 
       mapping = aes(x = lotsize)) +
  geom_histogram(fill = "darkgreen", color = "red") +
  labs(title = "Most homes have a lot size between 5,000 and 7,500", 
       x = "Lot size", y = "Count")
```


(3) Develop an effective visualization on your own using the code chunk provided
    below. Use three variables and at least one aesthetic mapping.

```{r on-your-own}
ggplot(mn_homes, mapping = aes(x = reorder(community, -salesprice, median), 
                               y = salesprice)) +
   geom_boxplot(fill = "purple", alpha = 0.5) +
   labs(x = "Community in Minneapolis", y = "Sale price ($)",
        title = "Why is Calhoun-Isles so expensive?")
```

Here I am using `reorder()` to sort the boxplots by their median values in
descending order. This allows for much easier comparisons compared to when
they are sorted alphabetically by community.


## Additional Resources

- https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-visualization-2.1.pdf
- https://github.com/GraphicsPrinciples/CheatSheet/blob/master/NVSCheatSheet.pdf
- https://ggplot2.tidyverse.org/
- http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html
- https://medium.com/bbc-visual-and-data-journalism/how-the-bbc-visual-and-data-journalism-team-works-with-graphics-in-r-ed0b35693535
- https://ggplot2-book.org/
- https://ggplot2.tidyverse.org/reference/geom_histogram.html
- https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
